Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-05-18 paper/letter.tex
some typos corrected

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MG$^3$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
features in its Lyman-$\alpha$ emission thad had evaded theoretical
interpretation so far. 
We show that these special features, a symmetric triple peaked
emission line, can be explained by gas bulk rotation.  
We find that the Lyman-$\alpha$ emission region in \tol\ should
have a rotational velocity of $V_{r}=300$ km s$^{-1}$ and a neutral
Hydrogen column density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Using archival observational information about \tol\ we find that the
diameter for that emission region diameter should be in the range of
$110$ pc $<D<$ $340$ pc and its total dynamical mass   should be
between $2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times
10^{9}$M$_{\odot}$.  
This dynamical mass is at least $16\pm 9$ times
larger than the neutral mass hydrogen.
We argue that a possibility to explain the excess in
dynamical mass is the presence of a super-massive black hole. }

The first generation of galaxies trace our cosmic origins. 
They were the first steps in the evolution of galaxies such as the Milky
Way. 
In the standard Big Bang cosmology the only elements that were
created in the nucleosynthesis process were Hydrogen, Helium and
Lithium. 
Heavier elemets must have been created in stellar evolution process. 
Therefore, we expect the first generation of
galaxies to be metal free and rich in Hydrogen. 
This kind of primordial galaxies have not been detected yet. 
However, dwarf star forming galaxies with a low metallicity content
are seen as templates to understand the early galaxy evolution process.

Almost fifty years ago \cite{PartridgePeebles} it was realized that
young galaxies could be detected through a strong Lyman-$\alpha$ line
emission.  
This theoretical prediction was only confirmed thirty year later on
distant relatively young, not primordial, galaxies.
Currently Lyman Alpha Emitting (LAE) galaxies are commonly targetted
in surveys. 
The presence of the Ly-$\alpha$ emission line provides confirmation of
the distance of a galaxy while provides clues about the stellar
population and inter-stellar medium conditions regulating the
Ly-$\alpha$ emission.

The Ly-$\alpha$ emission line is not exclusive of distant galaxies. 
Any galaxy with low dust content and ongoing star formation has the
right conditions to show this line.  
There are, for instance,  local Universe surveys that target
Ly-$\alpha$ emission in nearby dwarf star forming galaxies. 
\cite{LARS}. 
The study of nearby LAE samples has allowd the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

However, the physical interpratation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a difussion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytical tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.

For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art montecarlo radiative transfer simulations.   
Recent advances in these computational models
\cite{GaravitoCamargo2014} have shown that galaxy rotation imprints an
effect on the Lyman-$\alpha$ line.
The most important consequence of rotation is that even for a
spherical gas distribution, the line morphology now depends on the
viewing angle respect to the rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity. 
When the rotational velocity is close to the half-line width of the
static line the line becomes single peaked, as it is observed in
\tol, a unique feature that other theoretical models find
difficult to reproduce without invoking complex gas distributions.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two important
features. 
First, it has symmetric profile  without any systemic displacement
from the expected line-center. 
Second, it shows a triple peaked profile. 
Commonly, the Ly-$\alpha$ line features a single or asymetric double peak. 
These two special features in \tol\ had evaded a natural physical
interpratation so far.

We find that a rotating spherical gas distribution provides a
satisfactory explanation of \tol's puzzling features.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol with the
overplot from our best fit model from the full radiative transfer
simulation.   
The parameters for the best fit are a rotational velocity of
$v_{max}=300$\kms, an optical depth $\tau=1\times10^7$ and temperature
of $T=1.5\times 10^{4}$K.
The optical depth and temperature can be translated into a column
density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Our model is also able to constrain the angle
between the rotation axis and the observational line-of-sight to
$\theta=65^{\circ}\pm$. 
To determine the uncertainties we perform a Markov Chain Monte
Carlo calculation using an approximate analytical solution for the
Ly-$\alpha$ spectrum including rotational effects.

@@

We consider three possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass can be constrained by measurements of the 
stellar continuum, this can only represent $xx$M$_{\odot}$ of the total
contribution. 
Considering a standard dark matter halo, it can only constribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

@@

The implications of our finding for the study of LAEs, including the
first generation of galaxies are manifold. 
First of all, it demonstrates the importance of including rotation as
theoretical feature to interpret observed spectra. 
Secondly, it provides a motivation to explore under what conditions
rotation dominated dynamics dominate over outflow
dominated dynamics. 
Finally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistend with future observational
kinematic maps, it would provide a new information to test and probe
theories on the coevolution of galaxies and black holes in the first
generation of galaxies.

@@

\tol receeding velocity is $7785\pm 50$km s$^{-1}$, which translates
into a distance of $106.6$ Mpc (Hubble constant 73 Mpc km$^{-1}$
s$^{1}$)
Its metallicity is $\sim Z_{\odot}/24$ \cite{Izotov04} as derived from optical
spectroscopy.

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MG$^3$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
features in its Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
We show that these special features, a symmetric triple peaked
emission line, can be explained by gas bulk rotation.  
We find that the Lyman-$\alpha$ emission region in \tol\ should
have a rotational velocity of $V_{r}=300$ km s$^{-1}$ and a neutral
Hydrogen column density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Using archival observational information about \tol\ we find that the
diameter for that emission region diameter should be in the range of
$110$ pc $<D<$ $340$ pc and its total dynamical mass   should be
between $2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times
10^{9}$M$_{\odot}$.  
This dynamical mass is at least $16\pm 9$ times
larger than the neutral mass hydrogen.
We argue that a possibility to explain the excess in
dynamical mass is the presence of a super-massive black hole. }

The first generation of galaxies trace our cosmic origins. 
They were the first steps in the evolution of galaxies such as the Milky
Way. 
In the standard Big Bang cosmology the only elements that were
created in the nucleosynthesis process were Hydrogen, Helium and
Lithium. 
Heavier elements must have been created in stellar evolution process. 
Therefore, we expect the first generation of
galaxies to be metal free and rich in Hydrogen. 
This kind of primordial galaxies have not been detected yet. 
However, dwarf star forming galaxies with a low metallicity content
are seen as templates to understand the early galaxy evolution process.

Almost fifty years ago \cite{PartridgePeebles} it was realized that
young galaxies could be detected through a strong Lyman-$\alpha$ line
emission.  
This theoretical prediction was only confirmed thirty year later on
distant relatively young, not primordial, galaxies.
Currently Lyman Alpha Emitting (LAE) galaxies are commonly targeted
in surveys. 
The presence of the Ly-$\alpha$ emission line provides confirmation of
the distance of a galaxy while provides clues about the stellar
population and inter-stellar medium conditions regulating the
Ly-$\alpha$ emission.

The Ly-$\alpha$ emission line is not exclusive of distant galaxies. 
Any galaxy with low dust content and ongoing star formation has the
right conditions to show this line.  
There are, for instance,  local Universe surveys that target
Ly-$\alpha$ emission in nearby dwarf star forming galaxies. 
\cite{LARS}. 
The study of nearby LAE samples has allowed the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

However, the physical interpretation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.

For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art Monte Carlo radiative transfer simulations.   
Recent advances in these computational models
\cite{GaravitoCamargo2014} have shown that galaxy rotation imprints an
effect on the Lyman-$\alpha$ line.
The most important consequence of rotation is that even for a
spherical gas distribution, the line morphology now depends on the
viewing angle respect to the rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity. 
When the rotational velocity is close to the half-line width of the
static line the line becomes single peaked, as it is observed in
\tol, a unique feature that other theoretical models find
difficult to reproduce without invoking complex gas distributions.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two important
features. 
First, it has symmetric profile  without any systemic displacement
from the expected line-center. 
Second, it shows a triple peaked profile. 
Commonly, the Ly-$\alpha$ line features a single or asymmetric double peak. 
These two special features in \tol\ had evaded a natural physical
interpretation so far.

We find that a rotating spherical gas distribution provides a
satisfactory explanation of \tol's puzzling features.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol with the
overplot from our best fit model from the full radiative transfer
simulation.   
The parameters for the best fit are a rotational velocity of
$v_{max}=300$\kms, an optical depth $\tau=1\times10^7$ and temperature
of $T=1.5\times 10^{4}$K.
The optical depth and temperature can be translated into a column
density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Our model is also able to constrain the angle
between the rotation axis and the observational line-of-sight to
$\theta=65^{\circ}\pm$. 
To determine the uncertainties we perform a Markov Chain Monte
Carlo calculation using an approximate analytic solution for the
Ly-$\alpha$ spectrum including rotational effects.

@@

We consider three possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass can be constrained by measurements of the 
stellar continuum, this can only represent $xx$M$_{\odot}$ of the total
contribution. 
Considering a standard dark matter halo, it can only contribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

@@

The implications of our finding for the study of LAEs, including the
first generation of galaxies are manifold. 
First of all, it demonstrates the importance of including rotation as
theoretical feature to interpret observed spectra. 
Secondly, it provides a motivation to explore under what conditions
rotation dominated dynamics dominate over outflow
dominated dynamics. 
Finally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it would provide a new information to test and probe
theories on the co-evolution of galaxies and black holes in the first
generation of galaxies.

@@

\tol receding velocity is $7785\pm 50$km s$^{-1}$, which translates
into a distance of $106.6$ Mpc (Hubble constant 73 Mpc km$^{-1}$
s$^{1}$)
Its metallicity is $\sim Z_{\odot}/24$ \cite{Izotov04} as derived from optical
spectroscopy.
